A family of accessibility measures derived from spatial interaction principles

### 2025-04-07 paper/manuscript.qmd
Singly-constrained opportunity

In contrast to unconstrained accessibility, imposing a constraint -in this case a total opportunity constraint- allows for the comparison of differences and ratios. Each value is effectively in units of physicians, with the impedance units already accounted for by $\kappa^T$.

Consider the highest decay scenario ($f_1(c_{ij})$), zone 1 again captures an intermediate amount of physicians (`r tot_acc$V_tot_i_1[1]`). However, we can say that this is out of the `r tot_acc$V_tot_i_1 |> sum()` in the region, which allows us also to deduce that zone 1 captures `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[2]` and `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[3]` times more than zone 2 and 3. Values for the lesser decay ($f_2(c_{ij})$) and lowest decay ($f_3(c_{ij})$) scenarios are calculated separately, with decay scenario values also summing to equal `r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region. Between zonal comparisons can be more interpretably made for these travel behaviour scenarios as well.

@@

@@

Effectively transposing $i$ and $j$, this case of the totally-constrained accessibility measure (@eq-total-constrained-market) is an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 and @vickermanAccessibilityAttractionPotential1974. The unconstrained form of market potential, effectively the $i$ $j$ transpose of $V^0_{ij}$, has been used in recent research to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road]. To formulate this case, the total constraint may instead be applied to the mass of the *population* at $i$ instead of the opportunities at $j$.

@@

$$
M^T_j = \sum_i M^T_{ij} = \hat K^T \sum_i W^{(1)}_if(c_{ij})
$$

The constraint that we impose in this case is the total market potential equals the total population $O$ in the region :

@@

However, unlike the totally-constrained opportunity case, $\kappa^S_{i}$ captures competition from the population. Again, let's consider the highest decay scenario $f_1(c_{ij})$. In this case, zone 1 no longer captures an intermediate amount of physicians as in the totally-constrained opportunity case: it now captures the fewest in the region i.e., `r opp_acc$V_opp_i_1[1]` at zone 1, `r opp_acc$V_opp_i_1[2]` at zone 2, and `r opp_acc$V_opp_i_1[3]` at zone 3. Why may zone 1 capture `r (opp_acc$V_opp_i_1[1]/opp_acc$V_opp_i_1 |> sum()) |> scales::percent()` of the physicians regionally while this same zone captures `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1 |> sum()) |> scales::percent()` in the total opportunity constrained case? This difference is due to the single-opportunity constraint $\kappa^S_{i}$'s role in $V_i^S$. The only inputs needed in the total-opportunity constraint $\kappa^T$ is the total number of opportunities in the region $D$ as well as the associated opportunities at each $j$ and travel impedance. In other words, opportunities are proportionally allocated to each $i$ based on the proportion of impedance-weighted opportunities at all $j$ for that $i$ to the total sum of impedance-weighted opportunities in the entire region. Opportunities are allocated to $i$s, regardless of the mass weights of origin. Whereas the single opportunity constraint $\kappa^S_{i}$ requires the population at $i$ as an input. In fact, $\kappa^S_{i}$ is the proportion of the sum of impedance-weighted population at an $i$ to the sum of impedance-weighted population for the entire region. Hence, since zone 1 has the lowest amount of people in the region, and is not located -travel cost effectively- to other opportunities, $V_{1}^S$ is lowest.

@@

@@

>>>>

In contrast to unconstrained accessibility, imposing a constraint -in this case a total opportunity constraint- allows for the comparison of differences and ratios between regions and across different travel behaviour scenarios as well. Each value is effectively in units of physicians, with the impedance units already accounted for by $\kappa^T$.

Consider the highest decay scenario ($f_1(c_{ij})$), zone 1 again captures an intermediate amount of physicians (`r tot_acc$V_tot_i_1[1]`). However, we can say that this is out of the `r tot_acc$V_tot_i_1 |> sum()` physicians in the region, which allows us also to deduce that zone 1 captures `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[2]` and `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[3]` times more than zone 2 and 3. Values for the lesser decay ($f_2(c_{ij})$) and lowest decay ($f_3(c_{ij})$) scenarios are calculated separately, with decay scenario values also summing to equal `r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region.

@@

<!-- rafa: This is a very important takeaway. But this is true for any metric using the constant, right? 66666666 -->

@@

By transposing $i$ and $j$, we have an  accessibility measure constrained by total population (@eq-total-constrained-market), which effectively gives us an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 and @vickermanAccessibilityAttractionPotential1974. The unconstrained form of market potential, effectively the $i$ $j$ transpose of $V^0_{ij}$, has been used in recent research to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road]. To formulate this case, the total constraint may instead be applied to the mass of the *population* at $i$ instead of the opportunities at $j$.

@@

$$
M^T_j = \sum_i M^T_{ij} = \hat K^T \sum_i W^{(1)}_if(c_{ij})
$$
This measure is simply a form of passive accessibility, and indicates that each destination can be reached by an x number of people. The constraint that we impose in this case in practice means that the total market potential equals the total population $O$ in the region:

@@

However, unlike the totally-constrained opportunity case, $\kappa^S_{i}$ captures competition from the population. Again, let's consider the highest decay scenario $f_1(c_{ij})$. In this case, zone 1 no longer captures an intermediate amount of physicians as in the totally-constrained opportunity case: it now captures the fewest in the region i.e., `r opp_acc$V_opp_i_1[1]` at zone 1, `r opp_acc$V_opp_i_1[2]` at zone 2, and `r opp_acc$V_opp_i_1[3]` at zone 3. Why may zone 1 capture `r (opp_acc$V_opp_i_1[1]/opp_acc$V_opp_i_1 |> sum()) |> scales::percent()` of the physicians regionally while this same zone captures `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1 |> sum()) |> scales::percent()` in the total opportunity constrained case? This difference is due to the single-opportunity constraint $\kappa^S_{i}$'s role in $V_i^S$. The only inputs needed in the total-opportunity constraint $\kappa^T$ is the total number of opportunities in the region $D$ as well as the associated opportunities at each $j$ and travel impedance. In other words, opportunities are proportionally allocated to each $i$ based on the proportion of impedance-weighted opportunities at all $j$ for that $i$ to the total sum of impedance-weighted opportunities in the entire region. Opportunities are allocated to $i$s, regardless of the mass weights of origin. Whereas the single opportunity constraint $\kappa^S_{i}$ requires the population at $i$ as an input. In fact, $\kappa^S_{i}$ is the proportion of the sum of impedance-weighted population at an $i$ to the sum of impedance-weighted population for the entire region. Hence, since zone 1 has the lowest number of people in the region, and is not located -travel cost effectively- to other opportunities, $V_{1}^S$ is lowest.

@@

<!-- rafa: I think we don't need the paragraph below -->

@@

<!-- rafa: in the last sentence, you clarify the difference between 'total opportunity constraint' Vs  'single opportunity constraint' . What is confusing here is how the single opportunity constraint compares to 'total population constraint' . Again, the idea of including a table would be very helpful. The table shoud include one indicator in each row. The columns could include (1) name of the indicator, (2) explanation in plain language of the constraint, (3) equation of the constraint, (4) equation of the indicator, (5) example of the interpretation of the indicator -->
